---
title: "Preparing ABCD data for IDEA-RS"
author: "Eileen Xu"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      collapsed: false
---

```{r include=F,}
library(tidyverse)
library(plyr)
library(summarytools)

knitr::opts_chunk$set(echo = T, message = F, results = "asis", warning = F)
```

## Variables from the environmental risk score for adolescent depression developed by [Rocha et al (2021)](https://doi.org/10.1016/j.jaac.2019.12.004) in Pelotas, Brazil.

Corresponding variables from ABCD baseline data (release 4.0)

| Pelotas Variable | ABCD Variable Description | ABCD Field Name |
|------------------|--------------------|----------------------------------|
| Sex | Sex | sex |
| Skin colour | Ethnic group/Race (parent report) | Culture_and_Environment/abcd_meim01 \> meim_ethnic_id |
| Childhood maltreatment | Traumatic events (parent report) | Mental_Health/abcd_ptsd01 \> total |
| School failure | School disengagement (youth report) | Culture_and_Environment/srpf01 \> 15 + 17 |
| Social isolation | Loneliness (parent report) | Mental_Health/abcd_cbcl01 \>cbcl_q12_p |
| Fights involvement | Gets in many fights (parent report) | Mental_Health/abcd_cbcl01 \> cbcl_q37_p |
| Ran away from home | Runs away from home (parent report) | Mental_Health/abcd_cbcl01 \> cbcl_q67_p |
| Drug use (any) | Substance use summary score (youth report) | Substance_Use/abcd_ysu02 \> add all "ever tried" measures |
| Relationship with mother | Caregiver 1 relationship (youth report) | Culture_and_Environment/crpbi01 \> total crpbi_parent1_y : parent5_y |
| Relationship with father | Caregiver 2 relationship (youth report) | Culture_and_Environment/crpbi01 \> total crpbi_caregiver12_y : caregiver16_y |
| Relationship between parents | Family conflict (youth report) | Culture_and_Environment/abcd_fes01 \> total |

### Ethnicity

```{r}
eth1 = readRDS("G://data/abcd/release4.0/iii.data/Culture_and_Environment/abcd_meim01.rds")
eth1 = eth1[,c(1,12, 13)]
eth1 = eth1[grepl("baseline_", eth1$eventname),]
freq(eth1$meim_ethnic_id_p, style = "rmarkdown", plain.ascii = F, round.digits = 1, display.labels = F)

eth1$white = ifelse(eth1$meim_ethnic_id_p == "White/Caucasian", "white", "nonwhite")
```

### Traumatic events

```{r}
trauma1 = readRDS("G://data/abcd/release4.0/iii.data/Mental_Health/abcd_ptsd01.rds")
trauma1[,6:22] = lapply(trauma1[,6:22], mapvalues, from = c("Yes", "No"), to=c(1, 0))
trauma1$trauma = rowSums(sapply(trauma1[,6:22], function(x) as.numeric(as.character(x))), na.rm=T)
trauma1 = trauma1[grepl("baseline_", trauma1$eventname),]
trauma1 = trauma1[,c(1, 23)]
hist(trauma1$trauma, breaks = 17, xlim=c(0, 5))

env = merge(eth1, trauma1, by = "src_subject_id", all=T)
```

### School disengagement

Recoded from 1 = NO!; 2 = no; 3 = yes; 4 = YES! to 0 = no, 1 = yes.

Q15: "usually, school bores me" and 
Q17: "getting good grades is not so important to me"

```{r}
school = readRDS("G://data/abcd/release4.0/iii.data/Culture_and_Environment/srpf01.rds")
school = school[grepl("baseline_", school$visit),]
school$school_15_y = ifelse(school$school_15_y == 3 | school$school_15_y == 4, 1, 0)
school$school_17_y = ifelse(school$school_17_y == 3| school$school_17_y == 4, 1, 0)
school$disengagement = rowSums(school[,16:17], na.rm = T)
school = school[,c(1, 20)]

env = merge(env, school, by = "src_subject_id", all=T)
```

### CBCL: loneliness, fights and running from home

```{r}
cbcl = readRDS("G://data/abcd/release4.0/iii.data/Mental_Health/abcd_cbcl01.rds")
cbcl = cbcl %>% select("src_subject_id", "cbcl_q12_p", "cbcl_q37_p", "cbcl_q67_p", "eventname")
names(cbcl)=c("src_subject_id", "loneliness", "fights", "runsaway", "eventname")
cbcl = cbcl[grepl("baseline_", cbcl$eventname),]
cbcl$loneliness01 = ifelse(cbcl$loneliness == "Not True", 0, 1)
cbcl$fights01 = ifelse(cbcl$fights == "Not True", 0 , 1)
cbcl$run01 = ifelse(cbcl$runsaway == "Not True", 0 , 1)
cbcl$eventname = NULL

env = merge(env, cbcl, by = "src_subject_id", all = T)
```

### Substance use

```{r}
drug = readRDS("G://data/abcd/release4.0/iii.data/Substance_Use/abcd_ysu02.rds")
subs = read.delim("druglist.txt")
drug = drug[,c(1, which(colnames(drug) %in% subs$Field.name == T))]
drug$everdrugs = rowSums(drug[,2:38], na.rm=T)
drug$everdrugs = ifelse(drug$everdrugs == 0, 0, 1)
drug = drug[,c(1, 39)]

env = merge(env, drug, by = "src_subject_id", all = T)
```

### Relationship with caregivers (low score = bad)

```{r}
caregivers = readRDS("G://data/abcd/release4.0/iii.data/Culture_and_Environment/crpbi01.rds")
caregivers = caregivers[grepl("baseline_", caregivers$eventname),]
caregivers$care1 = rowSums(caregivers[, 5:9], na.rm=T)
caregivers$care2 = ifelse(caregivers$crpbi_caregiver1_y==1, rowSums(caregivers[, 12:16], na.rm=T), NA)
caregivers = caregivers[, c(1, 20, 21)]
hist(caregivers$care1)
hist(caregivers$care2)

env = merge(env, caregivers, by = "src_subject_id", all = T)
```

### Family conflict (low score = good)

```{r}
fam = readRDS("G://data/abcd/release4.0/iii.data/Culture_and_Environment/abcd_fes01.rds")
fam = fam[grepl("baseline_", fam$eventname),]
fam$conflict = rowSums(fam[,5:13], na.rm=T)
fam = fam[,c(1, 16)]

env = merge(env, fam, by = "src_subject_id", all = T)
```

### Separate caregiver relationships and conflict into quintiles

```{r}
# care 1 and care 2 are coded so that 1 = best, 5 = poorest
env$care1q = factor(ntile(x = env$care1, 5))
env$care1q = mapvalues(env$ care1q, from = c(1, 2, 3, 4, 5), 
                       to = c("poorest", "poor", "good","very good", "best"))
env$care2q = factor(ntile(x=env$care2, 5))
env$care2q = mapvalues(env$care2q, from = c(1, 2, 3, 4, 5), 
                       to = c("poorest", "poor", "good", "very good", "best"))

env$conflictq = as.factor(ntile(x=env$conflict, 5))
env$conflictq = mapvalues(env$conflictq, from = c(1, 2, 3, 4, 5),
                          to = c("least conflict", "little conflict", 
                                 "moderate conflict", "much conflict",
                                 "most conflict"))

env$trauma3 = as.factor(ifelse(env$trauma == 0, "none", ifelse(env$trauma == 1, "one", "multiple")))
env$sch1 = as.factor(ifelse(env$disengagement == 2, 1, 0))
```

## Manually coding MDD status and creating depression symptoms 

File to generate the following measures of depression: \* DSM-V diagnoses of Major Depressive Disorder, including impairment due to depression. \* Current depressive symptoms, mild/moderate/severe. Based on DSM-V symptoms and field names described in [Shen et. al. (2021)](https://doi.org/10.1016/j.eclinm.2021.101204)

The process was the same for parent-reported MDD, I've just included the youth-report here. I originally had these in separate scripts, but this is where the age and sex variables were extracted too.

### Baseline

Generate DSM-V diagnoses (at least 1 core symptom, at least five total symptoms and impairment) and measures of DS severity: \* 0 = control \* 1 = mild: at least 1 core symptom and \>1 secondary symptoms OR 5-6 total symptoms \* 2 = moderate: at least 1 core symptom and \>2 secondary symptoms OR 7-8 total symptoms \* 3 = severe: both core symptoms and \>3 secondary symptoms OR identified as clinical case and suicidality symptoms present

#### Current symptoms

```{r}
mdd.y = readRDS("G://data/abcd/release4.0/iii.data/Mental_Health/abcd_ksad501.rds")

mdd.y = data.frame(mdd.y[,1:6], mdd.y[, grepl("ksads_1_|ksads_23_|ksads_22", names(mdd.y))])
mdd.y = as.data.frame(sapply(mdd.y, gsub, pattern="not administered|not administered due to gating", replacement=NA))

mdd.y.base = mdd.y[grepl("baseline_", mdd.y$eventname),]

#RENAME COLUMNS TO THE CORRESPONDING DEPRESSION SYMPTOM#
DSlist = read.delim("DSlist.txt", stringsAsFactors = F) #list of ksads field names and corresponding depression symptoms

namelist = c(NULL) #create empty vector to fill new column name

###CURRENT SYMPTOMS###
for (i in 1:ncol(mdd.y.base)) {
  old.name = as.character(colnames(mdd.y.base)[i])
  new.name = if (old.name %in% DSlist$Field.name.current.1) {
    
  as.character(DSlist$Description[match(old.name, DSlist$Field.name.current.1)])
  } else  {old.name}
  namelist = append(namelist, new.name)
}

colnames(mdd.y.base) = namelist

#MAKE GROUPED VARIABLE OF WEIGHT/APPETITE CHANGE (EITHER GAIN OR LOSS)
weightsymptom = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    weightloss = mdd.y.base$`Weight Loss`[i]
    weightgain = mdd.y.base$`Weight Gain`[i]
    apploss = mdd.y.base$`Decreased Appetite`[i]
    appgain = mdd.y.base$`Increased Appetite`[i]
    change = if (is.na(weightloss)==T) {NA}
    else  if (weightloss == 1 | weightgain == 1 | apploss ==1 | appgain == 1) {1} else {weightloss}
    weightsymptom = append(weightsymptom, change)
}
mdd.y.base$WeightChange = weightsymptom
#MAKE VARIABLE OF SLEEP DISTURBANCE (EITHER INSOMNIA OR HYPERSOMNIA)
sleepsymptom = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    insomn = mdd.y.base$`Insomnia when depressed`[i]
    hypersomn = mdd.y.base$Hypersomnia[i]
    sleepdisturb = if (is.na(insomn)==T | is.na(hypersomn)==T) {NA}
    else  if (insomn == 1 | hypersomn == 1) {1} else {insomn}
    sleepsymptom = append(sleepsymptom, sleepdisturb)
}
mdd.y.base$SleepDisturb = sleepsymptom
#MAKE VARIABLE OF PSYCHOMOTOR DISTURBANCE
psychomotor = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    decrease = mdd.y.base$`Psychomotor Retardation`[i]
    increase = mdd.y.base$`Psychomotor Agitation in Depressive Disorder`[i]
    change = if (is.na(decrease)==T | is.na(increase)==T) {NA}
    else  if (decrease == 1 | increase == 1) {1} else {decrease}
    psychomotor = append(psychomotor, change)
}
mdd.y.base$PsychDisturb = psychomotor
#MAKE VARIABLE OF WORTHLESSNESS/GUILT
worthlessness = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    guilt = mdd.y.base$Guilt[i]
    self = mdd.y.base$`Decreased Self-Esteem`[i]
    change = if (is.na(guilt)==T | is.na(self)==T) {NA}
    else  if (guilt == 1 | self == 1) {1} else {guilt}
    worthlessness = append(worthlessness, change)
}
mdd.y.base$Worthlessness = worthlessness
#MAKE VARIABLE OF CONCENTRATION/INDECISIVENESS  
concentrate = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    disturb = mdd.y.base$`Concentration Disturbance`[i]
    indecision = mdd.y.base$Indecision[i]
    change = if (is.na(disturb)==T | is.na(indecision)==T) {NA}
    else  if (disturb == 1 | indecision == 1) {1} else {disturb}
    concentrate = append(concentrate, change)
}
mdd.y.base$Concentrate = concentrate
#MAKE VARIABLE OF DEPRESSED MOOD SYMPTOM: COUNT DEPRESSED MOOD, HOPELESSNESS AND IRRITABILITY AS ONE SYMPTOM
mood = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    depressed = mdd.y.base$`Depressed Mood`[i]
    hopeless = mdd.y.base$Hopeless[i]
    irritable = mdd.y.base$Irritability[i]
    change = if (is.na(depressed)==T | is.na(hopeless)==T | is.na(irritable)==T) {NA}
    else  if (depressed == 1 | hopeless == 1 | irritable == 1) {1} else {depressed}
    mood = append(mood, change)
}
mdd.y.base$MoodSymptom = mood

rm(appgain, apploss, change, concentrate, decrease, disturb, guilt, hypersomn, i, increase, indecision, insomn, namelist, new.name, old.name, psychomotor, self, sleepdisturb, sleepsymptom, weightgain, weightloss, weightsymptom, worthlessness,depressed, hopeless, irritable, mood)
```

```{r}
#CALCULATE TOTAL SCORES: CORE, SECONDARY AND SUICIDALITY
CoreSymptoms = "MoodSymptom|Anhedonia"
listcolsCore = grep(CoreSymptoms, colnames(mdd.y.base))
Suicidality = "Suicid|Attempt"
listcolsSui = grep(Suicidality, colnames(mdd.y.base))
SecSymptoms = "Fatigue|Concentrate|Worthlessness|SleepDisturb|WeightChange|PsychDisturb"
listcolsSec = grep(SecSymptoms, colnames(mdd.y.base))

mdd.y.base$CoreDSPresent = rowSums(sapply(mdd.y.base[,listcolsCore], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.base$SuicidalityPresent = rowSums(sapply(mdd.y.base[,listcolsSui], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.base$SecDSPresent = rowSums(sapply(mdd.y.base[,listcolsSec], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.base$TotalSymptoms = mdd.y.base$CoreDSPresent + mdd.y.base$SecDSPresent


#IDENTIFY CASES BY DSM-V DIAGNOSIS
mdd.y.base = mutate(mdd.y.base, DiagnosisDSM = ifelse(CoreDSPresent>=1 & TotalSymptoms>=5 & `Impairment in functioning due to depression`== 1 | CoreDSPresent>=1 & TotalSymptoms >=4 & SuicidalityPresent >0 & SuicidalityPresent<10 & `Impairment in functioning due to depression`== 1, 1, 0))

#CODE DEPRESSION SYMPTOMS BASED ON SEVERITY
mdd.y.base = mutate(mdd.y.base, Severity = ifelse(((CoreDSPresent==2 & SecDSPresent>3)|(SuicidalityPresent>=1))& DiagnosisDSM==1, 3, 
                                           ifelse((CoreDSPresent>1 & SecDSPresent>2)|(TotalSymptoms>=7 & TotalSymptoms<=8),2,
                                           ifelse((CoreDSPresent>1 & SecDSPresent>=1)|((TotalSymptoms>=5 & TotalSymptoms<=6)),1,0))))

rm(CoreSymptoms, listcolsCore, listcolsSec, listcolsSui, SecSymptoms, Suicidality)
```

#### Lifetime symptoms

For lifetime symptoms and diagnoses: rename past symptoms, adding ".P" prefix

```{r}
###PAST SYMPTOMS###
namelist = c(NULL)

for (i in 1:ncol(mdd.y.base)) {
  old.name = as.character(colnames(mdd.y.base)[i])
  new.name = if (old.name %in% DSlist$Field.name.past.1) {
  as.character(paste0("P.",DSlist$Description[match(old.name, DSlist$Field.name.past.1)]))
  } else  {old.name}
  namelist = append(namelist, new.name)
}

colnames(mdd.y.base) = namelist

#MAKE GROUPED VARIABLE OF WEIGHT/APPETITE CHANGE (EITHER GAIN OR LOSS)
weightsymptom = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    weightloss = mdd.y.base$`P.Weight Loss`[i]
    weightgain = mdd.y.base$`P.Weight Gain`[i]
    apploss = mdd.y.base$`P.Decreased Appetite`[i]
    appgain = mdd.y.base$`P.Increased Appetite`[i]
    change = if (is.na(weightloss)==T) {NA}
    else  if (weightloss == 1 | weightgain == 1 | apploss ==1 | appgain == 1) {1} else {weightloss}
    weightsymptom = append(weightsymptom, change)
}
mdd.y.base$P.WeightChange = weightsymptom
#MAKE VARIABLE OF SLEEP DISTURBANCE (EITHER INSOMNIA OR HYPERSOMNIA)
sleepsymptom = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    insomn = mdd.y.base$`P.Insomnia when depressed`[i]
    hypersomn = mdd.y.base$P.Hypersomnia[i]
    sleepdisturb = if (is.na(insomn)==T | is.na(hypersomn)==T) {NA}
    else  if (insomn == 1 | hypersomn == 1) {1} else {insomn}
    sleepsymptom = append(sleepsymptom, sleepdisturb)
}
mdd.y.base$P.SleepDisturb = sleepsymptom
#MAKE VARIABLE OF PSYCHOMOTOR DISTURBANCE
psychomotor = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    decrease = mdd.y.base$`P.Psychomotor Retardation`[i]
    increase = mdd.y.base$`P.Psychomotor Agitation in Depressive Disorder`[i]
    change = if (is.na(decrease)==T | is.na(increase)==T) {NA}
    else  if (decrease == 1 | increase == 1) {1} else {decrease}
    psychomotor = append(psychomotor, change)
}
mdd.y.base$P.PsychDisturb = psychomotor
#MAKE VARIABLE OF WORTHLESSNESS/GUILT
worthlessness = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    guilt = mdd.y.base$P.Guilt[i]
    self = mdd.y.base$`P.Decreased Self-Esteem`[i]
    change = if (is.na(guilt)==T | is.na(self)==T) {NA}
    else  if (guilt == 1 | self == 1) {1} else {guilt}
    worthlessness = append(worthlessness, change)
}
mdd.y.base$P.Worthlessness = worthlessness
#MAKE VARIABLE OF CONCENTRATION/INDECISIVENESS  
concentrate = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    disturb = mdd.y.base$`P.Concentration Disturbance`[i]
    indecision = mdd.y.base$P.Indecision[i]
    change = if (is.na(disturb)==T | is.na(indecision)==T) {NA}
    else  if (disturb == 1 | indecision == 1) {1} else {disturb}
    concentrate = append(concentrate, change)
}
mdd.y.base$P.Concentrate = concentrate
#MAKE VARIABLE OF DEPRESSED MOOD SYMPTOM: COUNT DEPRESSED MOOD, HOPELESSNESS AND IRRITABILITY AS ONE SYMPTOM
mood = c(NULL)
for (i in 1:nrow(mdd.y.base)) {
    depressed = mdd.y.base$`P.Depressed Mood`[i]
    hopeless = mdd.y.base$P.Hopeless[i]
    irritable = mdd.y.base$P.Irritability[i]
    change = if (is.na(depressed)==T | is.na(hopeless)==T | is.na(irritable)==T) {NA}
    else  if (depressed == 1 | hopeless == 1 | irritable == 1) {1} else {depressed}
    mood = append(mood, change)
}
mdd.y.base$P.MoodSymptom = mood

rm(appgain, apploss, change, concentrate, decrease, disturb, guilt, hypersomn, i, increase, indecision, insomn, namelist, new.name, old.name, psychomotor, self, sleepdisturb, sleepsymptom, weightgain, weightloss, weightsymptom, worthlessness, depressed, hopeless, irritable, mood)
```

```{r}
#CALCULATE TOTAL SCORES: CORE, SECONDARY AND SUICIDALITY
CoreSymptoms = "P.MoodSymptom|P.Anhedonia"
listcolsCore = grep(CoreSymptoms, colnames(mdd.y.base))
Suicidality = "P.Suicid|P.Attempt"
listcolsSui = grep(Suicidality, colnames(mdd.y.base))
SecSymptoms = "P.Fatigue|P.Concentrate|P.Worthlessness|P.SleepDisturb|P.WeightChange|P.PsychDisturb"
listcolsSec = grep(SecSymptoms, colnames(mdd.y.base))


mdd.y.base$CoreDSPast = rowSums(sapply(mdd.y.base[,listcolsCore], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.base$SuicidalityPast = rowSums(sapply(mdd.y.base[,listcolsSui], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.base$SecDSPast = rowSums(sapply(mdd.y.base[,listcolsSec], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.base$TotalSymptomsPast = mdd.y.base$CoreDSPresent + mdd.y.base$SecDSPresent


#IDENTIFY CASES BY DSM-V DIAGNOSIS
mdd.y.base = mutate(mdd.y.base, DiagnosisDSMPast = ifelse(CoreDSPast>=1 & TotalSymptomsPast>=5 & `P.Impairment in functioning due to depression`== 1 | CoreDSPast>=1 & TotalSymptomsPast >=4 & SuicidalityPast >0 & SuicidalityPast<10 & `P.Impairment in functioning due to depression`== 1, 1, 0))

#CODE DEPRESSION SYMPTOMS BASED ON SEVERITY
mdd.y.base = mutate(mdd.y.base, SeverityPast = ifelse(((CoreDSPast==2 & SecDSPast>3)|(SuicidalityPast>=1))& DiagnosisDSMPast==1, 3, 
                                           ifelse((CoreDSPast>1 & SecDSPast>2)|(TotalSymptomsPast>=7 & TotalSymptomsPast<=8),2,
                                           ifelse((CoreDSPast>1 & SecDSPast>=1)|((TotalSymptomsPast>=5 & TotalSymptomsPast<=6)),1,0))))

rm(CoreSymptoms, listcolsCore, listcolsSec, listcolsSui, SecSymptoms, Suicidality)

#COMBINE INTO LIFETIME SCORE

mdd.y.base = mutate(mdd.y.base, DiagnosisDSMLifetime = ifelse(DiagnosisDSM ==1 | DiagnosisDSMPast ==1, 1, 0))
mdd.y.base = mutate(mdd.y.base, SeverityLifetime = ifelse(Severity == 3 | SeverityPast ==3, 3,
                                                          ifelse(Severity == 2 | SeverityPast ==2, 2,
                                                                 ifelse(Severity ==1 | SeverityPast ==1, 1, 0))))

```

```{r}
#Keep relevant columns only
mdd.y.base = mdd.y.base %>% select(1:6, 113:118, 125:132)
#saveRDS(mdd.y.base, "...")
MDDbase = mdd.y.base[,c(1,3,4,11,12,17:20)]
colnames(MDDbase) = c("src_subject_id", "age","sex", "MDD_b", "Severity_b", "MDDPast_b", "SeverityPast_b", "MDDLifetime_b", "SeverityLifetime_b")
```

### Year 2

Repeat the above for youth-reported depression at 2-year follow-up

#### Current symptoms 

```{r}
mdd.y.2yr = mdd.y[grepl("2_year", mdd.y$eventname),]

#RENAME COLUMNS TO THE CORRESPONDING DEPRESSION SYMPTOM#
DSlist = read.delim("DSlist.txt", stringsAsFactors = F) #list of ksads field names and corresponding depression symptoms

namelist = c(NULL) #create empty vector to fill new column name

###CURRENT SYMPTOMS###
for (i in 1:ncol(mdd.y.2yr)) {
  old.name = as.character(colnames(mdd.y.2yr)[i])
  new.name = if (old.name %in% DSlist$Field.name.current.1) {
  as.character(DSlist$Description[match(old.name, DSlist$Field.name.current.1)])
  } else  {old.name}
  namelist = append(namelist, new.name)
}

colnames(mdd.y.2yr) = namelist

#MAKE GROUPED VARIABLE OF WEIGHT/APPETITE CHANGE (EITHER GAIN OR LOSS)
weightsymptom = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    weightloss = mdd.y.2yr$`Weight Loss`[i]
    weightgain = mdd.y.2yr$`Weight Gain`[i]
    apploss = mdd.y.2yr$`Decreased Appetite`[i]
    appgain = mdd.y.2yr$`Increased Appetite`[i]
    change = if (is.na(weightloss)==T) {NA}
    else  if (weightloss == 1 | weightgain == 1 | apploss ==1 | appgain == 1) {1} else {weightloss}
    weightsymptom = append(weightsymptom, change)
}
mdd.y.2yr$WeightChange = weightsymptom
#MAKE VARIABLE OF SLEEP DISTURBANCE (EITHER INSOMNIA OR HYPERSOMNIA)
sleepsymptom = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    insomn = mdd.y.2yr$`Insomnia when depressed`[i]
    hypersomn = mdd.y.2yr$Hypersomnia[i]
    sleepdisturb = if (is.na(insomn)==T | is.na(hypersomn)==T) {NA}
    else  if (insomn == 1 | hypersomn == 1) {1} else {insomn}
    sleepsymptom = append(sleepsymptom, sleepdisturb)
}
mdd.y.2yr$SleepDisturb = sleepsymptom
#MAKE VARIABLE OF PSYCHOMOTOR DISTURBANCE
psychomotor = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    decrease = mdd.y.2yr$`Psychomotor Retardation`[i]
    increase = mdd.y.2yr$`Psychomotor Agitation in Depressive Disorder`[i]
    change = if (is.na(decrease)==T | is.na(increase)==T) {NA}
    else  if (decrease == 1 | increase == 1) {1} else {decrease}
    psychomotor = append(psychomotor, change)
}
mdd.y.2yr$PsychDisturb = psychomotor
#MAKE VARIABLE OF WORTHLESSNESS/GUILT
worthlessness = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    guilt = mdd.y.2yr$Guilt[i]
    self = mdd.y.2yr$`Decreased Self-Esteem`[i]
    change = if (is.na(guilt)==T | is.na(self)==T) {NA}
    else  if (guilt == 1 | self == 1) {1} else {guilt}
    worthlessness = append(worthlessness, change)
}
mdd.y.2yr$Worthlessness = worthlessness
#MAKE VARIABLE OF CONCENTRATION/INDECISIVENESS  
concentrate = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    disturb = mdd.y.2yr$`Concentration Disturbance`[i]
    indecision = mdd.y.2yr$Indecision[i]
    change = if (is.na(disturb)==T | is.na(indecision)==T) {NA}
    else  if (disturb == 1 | indecision == 1) {1} else {disturb}
    concentrate = append(concentrate, change)
}
mdd.y.2yr$Concentrate = concentrate
#MAKE VARIABLE OF DEPRESSED MOOD SYMPTOM: COUNT DEPRESSED MOOD, HOPELESSNESS AND IRRITABILITY AS ONE SYMPTOM
mood = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    depressed = mdd.y.2yr$`Depressed Mood`[i]
    hopeless = mdd.y.2yr$Hopeless[i]
    irritable = mdd.y.2yr$Irritability[i]
    change = if (is.na(depressed)==T | is.na(hopeless)==T | is.na(irritable)==T) {NA}
    else  if (depressed == 1 | hopeless == 1 | irritable == 1) {1} else {depressed}
    mood = append(mood, change)
}
mdd.y.2yr$MoodSymptom = mood

rm(appgain, apploss, change, concentrate, decrease, disturb, guilt, hypersomn, i, increase, indecision, insomn, namelist, new.name, old.name, psychomotor, self, sleepdisturb, sleepsymptom, weightgain, weightloss, weightsymptom, worthlessness, depressed, hopeless, irritable, mood)
```

```{r}
#CALCULATE TOTAL SCORES: CORE, SECONDARY AND SUICIDALITY
CoreSymptoms = "MoodSymptom|Anhedonia"
listcolsCore = grep(CoreSymptoms, colnames(mdd.y.2yr))
Suicidality = "Suicid|Attempt"
listcolsSui = grep(Suicidality, colnames(mdd.y.2yr))
SecSymptoms = "Fatigue|Concentrate|Worthlessness|SleepDisturb|WeightChange|PsychDisturb"
listcolsSec = grep(SecSymptoms, colnames(mdd.y.2yr))


mdd.y.2yr$CoreDSPresent = rowSums(sapply(mdd.y.2yr[,listcolsCore], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.2yr$SuicidalityPresent = rowSums(sapply(mdd.y.2yr[,listcolsSui], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.2yr$SecDSPresent = rowSums(sapply(mdd.y.2yr[,listcolsSec], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.2yr$TotalSymptoms = mdd.y.2yr$CoreDSPresent + mdd.y.2yr$SecDSPresent

#IDENTIFY CASES BY DSM-V DIAGNOSIS
mdd.y.2yr = mutate(mdd.y.2yr, DiagnosisDSM = ifelse(CoreDSPresent>=1 & TotalSymptoms>=5 & `Impairment in functioning due to depression`== 1 | CoreDSPresent>=1 & TotalSymptoms >=4 & SuicidalityPresent >0 & SuicidalityPresent<10 & `Impairment in functioning due to depression`== 1, 1, 0))

#CODE DEPRESSION SYMPTOMS BASED ON SEVERITY
mdd.y.2yr = mutate(mdd.y.2yr, Severity = ifelse(((CoreDSPresent==2 & SecDSPresent>3)|(SuicidalityPresent>=1))& DiagnosisDSM==1, 3, 
                                           ifelse((CoreDSPresent>1 & SecDSPresent>2)|(TotalSymptoms>=7 & TotalSymptoms<=8),2,
                                           ifelse((CoreDSPresent>1 & SecDSPresent>=1)|((TotalSymptoms>=5 & TotalSymptoms<=6)),1,0))))

rm(CoreSymptoms, listcolsCore, listcolsSec, listcolsSui, SecSymptoms, Suicidality)
```

#### Lifetime symptoms

```{r}
###PAST SYMPTOMS###
namelist = c(NULL)

for (i in 1:ncol(mdd.y.2yr)) {
  old.name = as.character(colnames(mdd.y.2yr)[i])
  new.name = if (old.name %in% DSlist$Field.name.past.1) {
  as.character(paste0("P.",DSlist$Description[match(old.name, DSlist$Field.name.past.1)]))
  } else  {old.name}
  namelist = append(namelist, new.name)
}

colnames(mdd.y.2yr) = namelist

#MAKE GROUPED VARIABLE OF WEIGHT/APPETITE CHANGE (EITHER GAIN OR LOSS)
weightsymptom = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    weightloss = mdd.y.2yr$`P.Weight Loss`[i]
    weightgain = mdd.y.2yr$`P.Weight Gain`[i]
    apploss = mdd.y.2yr$`P.Decreased Appetite`[i]
    appgain = mdd.y.2yr$`P.Increased Appetite`[i]
    change = if (is.na(weightloss)==T) {NA}
    else  if (weightloss == 1 | weightgain == 1 | apploss ==1 | appgain == 1) {1} else {weightloss}
    weightsymptom = append(weightsymptom, change)
}
mdd.y.2yr$P.WeightChange = weightsymptom
#MAKE VARIABLE OF SLEEP DISTURBANCE (EITHER INSOMNIA OR HYPERSOMNIA)
sleepsymptom = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    insomn = mdd.y.2yr$`P.Insomnia when depressed`[i]
    hypersomn = mdd.y.2yr$P.Hypersomnia[i]
    sleepdisturb = if (is.na(insomn)==T | is.na(hypersomn)==T) {NA}
    else  if (insomn == 1 | hypersomn == 1) {1} else {insomn}
    sleepsymptom = append(sleepsymptom, sleepdisturb)
}
mdd.y.2yr$P.SleepDisturb = sleepsymptom
#MAKE VARIABLE OF PSYCHOMOTOR DISTURBANCE
psychomotor = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    decrease = mdd.y.2yr$`P.Psychomotor Retardation`[i]
    increase = mdd.y.2yr$`P.Psychomotor Agitation in Depressive Disorder`[i]
    change = if (is.na(decrease)==T | is.na(increase)==T) {NA}
    else  if (decrease == 1 | increase == 1) {1} else {decrease}
    psychomotor = append(psychomotor, change)
}
mdd.y.2yr$P.PsychDisturb = psychomotor
#MAKE VARIABLE OF WORTHLESSNESS/GUILT
worthlessness = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    guilt = mdd.y.2yr$P.Guilt[i]
    self = mdd.y.2yr$`P.Decreased Self-Esteem`[i]
    change = if (is.na(guilt)==T | is.na(self)==T) {NA}
    else  if (guilt == 1 | self == 1) {1} else {guilt}
    worthlessness = append(worthlessness, change)
}
mdd.y.2yr$P.Worthlessness = worthlessness
#MAKE VARIABLE OF CONCENTRATION/INDECISIVENESS  
concentrate = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    disturb = mdd.y.2yr$`P.Concentration Disturbance`[i]
    indecision = mdd.y.2yr$P.Indecision[i]
    change = if (is.na(disturb)==T | is.na(indecision)==T) {NA}
    else  if (disturb == 1 | indecision == 1) {1} else {disturb}
    concentrate = append(concentrate, change)
}
mdd.y.2yr$P.Concentrate = concentrate
#MAKE VARIABLE OF DEPRESSED MOOD SYMPTOM: COUNT DEPRESSED MOOD, HOPELESSNESS AND IRRITABILITY AS ONE SYMPTOM
mood = c(NULL)
for (i in 1:nrow(mdd.y.2yr)) {
    depressed = mdd.y.2yr$`P.Depressed Mood`[i]
    hopeless = mdd.y.2yr$P.Hopeless[i]
    irritable = mdd.y.2yr$P.Irritability[i]
    change = if (is.na(depressed)==T | is.na(hopeless)==T | is.na(irritable)==T) {NA}
    else  if (depressed == 1 | hopeless == 1 | irritable == 1) {1} else {depressed}
    mood = append(mood, change)
}
mdd.y.2yr$P.MoodSymptom = mood

rm(appgain, apploss, change, concentrate, decrease, disturb, guilt, hypersomn, i, increase, indecision, insomn, namelist, new.name, old.name, psychomotor, self, sleepdisturb, sleepsymptom, weightgain, weightloss, weightsymptom, worthlessness, depressed, hopeless, irritable, mood)
```

```{r}
#CALCULATE TOTAL SCORES: CORE, SECONDARY AND SUICIDALITY
CoreSymptoms = "P.MoodSymptom|P.Anhedonia"
listcolsCore = grep(CoreSymptoms, colnames(mdd.y.2yr))
Suicidality = "P.Suicid|P.Attempt"
listcolsSui = grep(Suicidality, colnames(mdd.y.2yr))
SecSymptoms = "P.Fatigue|P.Concentrate|P.Worthlessness|P.SleepDisturb|P.WeightChange|P.PsychDisturb"
listcolsSec = grep(SecSymptoms, colnames(mdd.y.2yr))

mdd.y.2yr$CoreDSPast = rowSums(sapply(mdd.y.2yr[,listcolsCore], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.2yr$SuicidalityPast = rowSums(sapply(mdd.y.2yr[,listcolsSui], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.2yr$SecDSPast = rowSums(sapply(mdd.y.2yr[,listcolsSec], function(x) as.numeric(as.character(x))), na.rm=T)
mdd.y.2yr$TotalSymptomsPast = mdd.y.2yr$CoreDSPresent + mdd.y.2yr$SecDSPresent


#IDENTIFY CASES BY DSM-V DIAGNOSIS
mdd.y.2yr = mutate(mdd.y.2yr, DiagnosisDSMPast = ifelse(CoreDSPast>=1 & TotalSymptomsPast>=5 & `P.Impairment in functioning due to depression`== 1 | CoreDSPast>=1 & TotalSymptomsPast >=4 & SuicidalityPast >0 & SuicidalityPast<10 & `P.Impairment in functioning due to depression`== 1, 1, 0))

#CODE DEPRESSION SYMPTOMS BASED ON SEVERITY
mdd.y.2yr = mutate(mdd.y.2yr, SeverityPast = ifelse(((CoreDSPast==2 & SecDSPast>3)|(SuicidalityPast>=1))& DiagnosisDSMPast==1, 3, 
                                           ifelse((CoreDSPast>1 & SecDSPast>2)|(TotalSymptomsPast>=7 & TotalSymptomsPast<=8),2,
                                           ifelse((CoreDSPast>1 & SecDSPast>=1)|((TotalSymptomsPast>=5 & TotalSymptomsPast<=6)),1,0))))

rm(CoreSymptoms, listcolsCore, listcolsSec, listcolsSui, SecSymptoms, Suicidality)

#COMBINE INTO LIFETIME SCORE

mdd.y.2yr = mutate(mdd.y.2yr, DiagnosisDSMLifetime = ifelse(DiagnosisDSM ==1 | DiagnosisDSMPast ==1, 1, 0))
mdd.y.2yr = mutate(mdd.y.2yr, SeverityLifetime = ifelse(Severity == 3 | SeverityPast ==3, 3,
                                                          ifelse(Severity == 2 | SeverityPast ==2, 2,
                                                                 ifelse(Severity ==1 | SeverityPast ==1, 1, 0))))

```

```{r}
#Keep relevant columns only
mdd.y.2yr = mdd.y.2yr %>% select(1:6, 113:118, 125:132)
#saveRDS(mdd.y.2yr, "...")
MDD2yr = mdd.y.2yr[,c(1,11,12,17:20)]
colnames(MDD2yr) = c("src_subject_id", "MDD_2y", "Severity_2y", "MDDPast_2y", "SeverityPast_2y", "MDDLifetime_2y", "SeverityLifetime_2y")
```

## Merge into final df

``` {r}
mdd = merge(MDDbase, MDD2yr, by = "src_subject_id", all = T)
dat = merge(mdd, env, by = "src_subject_id", all=T)
dat$age = as.numeric(dat$age)

fct_cols = c("sex", "white", "disengagement", "everdrugs", "loneliness01", "fights01", "run01", "care1q", "care2q", "conflictq", "trauma3", "sch1")

dat[,which(colnames(dat) %in% fct_cols)] = lapply(dat[,which(colnames(dat) %in% fct_cols)], function(x) as.factor(x))
```

### Make cols for incident MDD and change in DS

```{r}
dat$incidentMDD = ifelse((dat$MDD_2y==1|dat$MDDLifetime_2y==1) & (dat$MDD_b==0 & dat$MDDLifetime_b==0), 1, ifelse(dat$MDD_b==1|dat$MDDLifetime_b==1, NA, 0))

dat$incidentMDD = as.factor(dat$incidentMDD)

dat$SeverityChange = (as.numeric(dat$Severity_2y) + as.numeric(dat$SeverityLifetime_2y)) - (as.numeric(dat$Severity_b) + as.numeric(dat$SeverityLifetime_b))

dat$incSeverity = as.factor(ifelse((dat$SeverityChange > 0)==T, 1, 0))
```

### Make sure DS are ordered none<mild<mod<severe (0<1<2<3)

```{r}
DSlevels = c(0,1,2,3)

dat$Severity_b = factor(dat$Severity_b, levels = DSlevels)
dat$SeverityPast_b = factor(dat$SeverityPast_b, levels = DSlevels)
dat$SeverityLifetime_b = factor(dat$SeverityLifetime_b, levels = DSlevels)
dat$Severity_2y = factor(dat$Severity_2y, levels = DSlevels)
dat$SeverityPast_2y = factor(dat$SeverityPast_2y, levels = DSlevels)
dat$SeverityLifetime_2y = factor(dat$SeverityLifetime_2y, levels = DSlevels)
```

### Set reference categories

```{r}
dat$sex = relevel(dat$sex, ref="M")
dat$white = relevel(dat$white, ref = "white")
dat$disengagement = relevel(dat$disengagement, ref = "0")
dat$everdrugs = relevel(dat$everdrugs, ref = "0")
dat$loneliness01 = relevel(dat$loneliness01, ref = "0")
dat$fights01 = relevel(dat$fights01, ref = "0")
dat$run01 = relevel(dat$run01, ref = "0")
dat$care1q = relevel(dat$care1q, ref = "best")
dat$care2q = relevel(dat$care2q, ref = "best")
dat$conflictq = relevel(dat$conflictq, ref = "least conflict")
dat$trauma3 = relevel(dat$trauma3, ref = "none")
dat$sch1 = relevel(dat$sch1, ref = "0")
```

``` {r results="asis"}
dfSummary(dat, graph.col = F)
#saveRDS(dat, "ABCD_ERS_youth.rds")
```

